Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-05-25 paper_jasa.tex
type I error

This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim B_{K, 1/m}$, where $B$ is a binomial distribution, therefore:

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ is given as 
\[
P(X \ge x) = 1 - \sum_{i=0}^{x-1} B_{K, 1/m} (i) = 1- \sum_{i=0}^{x-1} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i},
\]
for $x > 0$, with $X$ defined as above.
\end{dfn}
\noindent Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the observed data plot.

\begin{table}[htp]
\begin{center}
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\\hline
1 &  1 & 0.0500 & 2 &  1 & 0.9050 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0049 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &        &   &   &        & 4 & 4 & 0.0000 & 5 & 4 & 0.0000 \\\hline
\end{tabular}
\end{center}
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m$ lineups.}
\label{pvalue}
\end{table}

@@

\begin{dfn}\label{dfn:error}
The \textbf{Type I error} is the probability of rejecting $H_0$ when it is true, the probability for that is $P(X \ge x_{1-\alpha})$, which is controlled by $\alpha$.
% identifying the observed data plot, when $H_0$ is true, which is $1/m$, and is thus controlled by the number of null plots. 
\textbf{Type II error} is the probability of failing to identify the observed data plot, when $H_0$ is not true, $P( X \le x_{1-\alpha})$, where $x_{1 - \alpha}$ is the (approximate) $1-\alpha$ quantile of  $B_{K, 1/m}$.
\end{dfn}

\green{Why are we including $x_{1-\alpha}$ in both sides?}

>>>>

This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim \text{Binom}({K, 1/m})$,  therefore:

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ is given as 
\[
P(X \ge x) = 1 - \text{Binom}({K, 1/m}) (x-1) = \sum_{i=x}^{K} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i},
\]
with $X$ defined as above. 
\end{dfn}
Note that for $x=0$ the $p$-value becomes --mathematically-- equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision.

\noindent Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the observed data plot.

\begin{table}[htp]
\begin{center}
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\\hline
1 &  1 & 0.0500 & 2 &  1 & 0.9050 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0049 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &        &   &   &        & 4 & 4 & 0.0000 & 5 & 4 & 0.0000 \\\hline
\end{tabular}
\end{center}
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\label{pvalue}
\end{table}

@@

\begin{dfn}\label{dfn:error}
The \textbf{Type I error} is the probability of rejecting $H_0$ when it is true, the probability for that is $P(X \ge x_{1-\alpha})$, which is controlled by $\alpha$.
% identifying the observed data plot, when $H_0$ is true, which is $1/m$, and is thus controlled by the number of null plots. 
\textbf{Type II error} is the probability of failing to identify the observed data plot, when $H_0$ is not true, $P( X <  x_{1-\alpha})$, where $x_{1 - \alpha}$ is the (approximate) $1-\alpha$ quantile of  $B_{K, 1/m}$.
\end{dfn}

\green{Why are we including $x_{1-\alpha}$ in both sides?}
\red{fixed  now-- $X < x_{1 - \alpha}$ is the acceptance region, $X \ge x_{1-\alpha}$ is rejection region.}
